### 2015-04-07 src/intro/intro.tex
Updated introduction

Because flagellar motion is limited to either tumbles – random reorientation movements – or runs, the pathway linking these flagella to the biochemical sensors consists of only a few steps. With the advent of more advanced sensory and motoric systems, both the amount of information available to drive behaviour as well as the amount of possible behaviours available increased tremendously. As direct biochemical links bewteen perception and action were no longer sufficient, simple neural network evolved to process the sensory signals allowing for abstract decisions about movement. As organisms became more complex, these networks evolved into a complete (central) nervous system, culminating in the cerebral cortex.

@@

In the following section, sensory sources for gravity and self-motion perception are considered in more detail. We then elaborate on how these signals may be used and integrated in the brain, and conclude the chapter with methods for studying the processing of these signals.

@@

\subsubsection{Otoliths}
The linear acceleration and gravitational forces are measured by the four otolith organs, each side of the head containing a saccule and an utricule. Each otolith consists of an endolymph filled compartment containing calcium carbonate crystals known as otoconia. Due to their high inertia, these otoconia fall behind during linear acceleration, and move ahead during deceleration. In addition, they fall downward as a result of the graviational field. The otoconia are mounted on top of a flexible polysaccharide gel, in which hair cells, similar to those found in the semicircular canals, are partly embedded \cite{goldberg2012}.

However, Einstein's equivalence principle states that  it is not possible to measure forces caused by linear acceleration and by gravity independently. The signal coming from the otoliths is therefore proportional to the combination of these forces \cite{fernandez1976b}, which is commonly referred to as the gravitoinertial force (GIF). Compared to the three semicircular canals, which are able to sense rotation in three dimensions due to their orthogonal organization, either side of the head only contains two otolith organs. It is still possible to sense the gravitoinertial force in three dimensions because the otolith organs are curved, and the orientation of the hair cells determined the direction of sensitivity \cite{goldberg2012}.

@@

For many actions, the brain needs to disentangle the contributions of gravity and linear acceleration to the gravitointertial force. For example, the linear vestibuloocular reflex (LVOR) that stabilizes gaze during translation should be sensitive to translation while ignoring gravity. Under normal circumstances, the brain is able to perform this task well \cite{merfeld1995}, but during extreme conditions such as in airplanes or space flight errors might occur. The somatogravic illusion \cite{glasauer1995} is an example of such error. In this illusion an airplane accelerates forward which causes an inertial force in the opposite direction, the resulting GIF will therefore be in between gravity (downward) and the intertial force (backward). The brain erroneously interprets the (majority of the) GIF as being caused by gravity, leading to the perception of "nose-up" pitch tilt.

Various theories have been proposed as to how the brain might solve the tilt-translation ambiguity. The frequency segregation hypothesis, for example, makes use of the fact that when stationary, the only force we experience is gravity. In this case, sustained (low frequency) accelerations should be  attributed to gravity, while the high frequency components should be attributed to translation \cite{paige1991, telford1997}. Other models keep track of the expected direction of gravity by mathematically integrating the rotational signal from the semi-circular canals, subtracting that signal from the otolith signal to obtain an estimate of translation. Because the semi-circular canals play a  crucial role in these models, they are known as multisensory integration models \cite{mayne1974,ormsby1977}. Merfeld and Zupan \citeyear{merfeld1995,merfeld2002} further refined the multisensory integration model by explicitly stating that the brain uses an (internal) model of the physical world to resolve the tilt-translation ambiguity.

The disambiguation of linear acceleration and gravity is not solely based on vestibular inputs, but also takes the  range of  possible movements into account. For example when seated on a linear sled, the probability of a participant perceiving tilt is greatly reduced \cite{wertheim2001}, indicating that cognitive processes also influence disambiguation.

@@

One complication is that the magnitude, $e$, of these eye movements not only depend on translation amplitude, $T$, but also on fixation depth, d. When fixating position $(x, d)$, the eye position at starting position is $e_0 = atan⁡(d/x) \approx d/x$. By subtracting the eye position after the translation, e_1=atan⁡(d/(x+T))≈d/(x+T), we obtain the eye movement amplitude, $e \approx d/T$. In Chapter 4 we explore whether the brain takes this geometry into account and compensates for fixation depth when using eye movements in self-motion perception.

\section{Reflexive uses for the self-motion}
While vection is an ideal cue in self-motion perception, it can cause the image on the fovea to blur which hampers visual perception. Several reflexive mechanisms attempt to keep fixation on target during self-motion. Two examples are the vestibulo-ocular reflex, the VOR, where the low-latency vestibular signal drives the reflex and the optokinetic reflex, the OKR, where optic flow acts as the source signal.

The optokinetic reflex [Expand on this]

As these VOR compensate for both rotation as well as translation, the VOR consists of two reflex arcs. The translational VOR, or TVOR, is driven by the otolith signal while the rotational VOR, or RVOR, is mainly driven by the semicircular canals. Unlike the visual system, the vestibular system does not require elaborate processing in order to drive the reflexive eye movements. Because of this, the reflex arc reaches the oculomotor system in just three neural links. This greatly reduces the latency of the VOR compared to that of the OKR and causes the VOR to dominate at higher frequencies \cite{schweigart1997}.

In both the OKR and the VOR the eye movements made consist of two parts, the smooth pursuit movement in which the eyes are kept on target using slow eye movements and the saccade in which the eyes are quickly moved back to their initial position. The fast movements are elicited when the eyes are about to lose track of the target because of physical constraints within the oculomotor system.

%Mergner and Rosemeier, 1998

@@

While reflexes depend on minimally processed sensory signals by definition, higher animal functions can rely on more elaborate processing. One such processing step is determining the underlying physical cause of a sensory signal. It has become clear that multiple sensory systems provide information about the same physical quantity. For example, both the visual and the vestibular system contribute to self-motion perception. A very naive approach to deal with these inout inputs would be to focus on the most reliable cue and ignore the others. A better solution, however, is to weight all available cues by their  relative reliabilities. This approach is known as statistically optimal, or Bayesian, integration, the latter referring to the use  of Bayes' inference.

@@

These reference frame transformations are not static, but depend on the position of our body. For example when converting the gaze-fixed retinal signal into a head-centered frame of reference, the position of the eyes have to be taken into account. When the position of our body changes, e.g. when we move our eyes, all coordinates remembered within the moving frame of reference also need to be updated. For example, when moving the eyes remembered retinal information needs to be shifted in the opposite direction to remain at the correct relative position. This process is known as spatial updating or spatial constancy.

@@

\subsection{Subjective body tilt}
In the subjective body tilt, or SBT, task the perceived body orientation with respect to a given body tilt angle is probed. Participants are first given a reference ange, e.g. 45 \si{\degree}, and are then rotated to an angle close to the reference angle, e.g. 46 \si{\degree} (see \figref{intro:fig4}A). They then have to indicate whether their current orientation is clockwise or counterclockwise with respect to the reference angle. By systematically probing rotation angles around the reference angle we can obtain both the bias and uncertainty on the percept of body tilt.

Most participants are able to do this perfectly, regardless of the reference angle (see \figref{intro:fig4}B). Their uncertainty does increase as a function of reference angle though (see \figref{intro:fig4}C).
As the SBT probes body orientation, the somatosensory signal originating from the torso can be used without any reference transformation and thus provides a direct contriubtion. Other sensory signals such as the vestibular signal can also be used, but only after a reference frame transformation.

@@

\subsection{Subjective visual vertical}
The subjective visual vertical, or SVV, task is a similar task in which participants have to judge the orientation of a line with respect to gravity. The PSE, that is the angle at which the line is perceived to be aligned with gravity, can be found by presenting lines at different angles and asking the participant whether the line is rotated clockwise or counterclockwise relative to gravity (see \figref{intro:fig5}A).

When seated straight this task, participants do not make any static errors and are very certain about their responses (see \figref{intro:fig5}B and C). This changes when first rotating the participant before the task. In general, the static error increases with tilt angle, the direction of the error is in the direction of the body midline, this effect is known as the Aubert or A-effect. At smaller angles overcompensation occurs and the static error moves in the oposite direction, that is away from the body midline. This latter effect is known as the E-effect and could be due to the effects of ocular counterroll (OCR).

@@

\section{Outline of this thesis}
Multiple sensory signals provide information about both body orientaiton as well as self-motion. This thesis explores the how the brain integrates these signals to form coherent percepts, and how it uses them to keep track of previously seen objects.

We start with Chapter 2 where we investigate whether the visual, somatosensory, and vestibular cues to body orientation are integrated in a statistically optimal fashion. Because studing the contributions of these signals in isolation is not possible, we adopted an inverse approach where we assumed statistical optimality and attempted to compute the statistical properties of individual sensors. To this end, we fitted an optimal integration model to two physophyisical tasks, one gauging head-in-space orientation (SVV) and the other body-in-space orientation (SBT). Predictions based on these noise properties were consistent with previously published deficits of various patient groups, further strengthening the idea that human spatial orientation is statistically optimal.

In Chapter 3 we move on to the dynamic aspects of vestibular integration, investigating whether eye movements contribute to the perception of whole-body translation. Using a psychophysical task, we first shows that translations in which the eyes fixate a body-fixed target are perceived as shorter than those in which the eyes fixate make pursuit movements by fixating a world-fixed target. We then go on an show that this result does not depend on the presence of the fixation point by including an unconstrained eye movement condition. Becase fixation distance influences the magnitude of pursuit eye movements, we fixed the distance of the fixation point in Chapter 3. To investigate whether the brain takes fixation distance into account, we compared self-motion perception during near and far world-fixed targets in Chapter 4. Our results suggest that translation is not properly accounted for, indicating that raw eye movements augment self-motion perception.

We conclude this thesis with Chapter 5, where we investigate how these self-motion signals are used to track previously seen objects across translation. First, we show that the errors made during spatial updating do not only depend on the location of the previously seen object, but also on the location of gaze. This is consistent with spatial updating in a gaze-centered reference frame. We further show that the underestimation of self-motion is a possible cause of the errors found.

>>>>

Because flagellar motion is limited to either tumbles – random reorientation movements – or runs, the pathway linking these flagella to the biochemical sensors consists of only a few steps. With the advent of more advanced sensory and motoric systems, both the amount of information available to drive behaviour as well as the amount of possible behaviours available increased tremendously. As direct biochemical links between perception and action were no longer sufficient, simple neural network evolved to process the sensory signals allowing for abstract decisions about movement. As organisms became more complex, these networks evolved into a complete (central) nervous system, culminating in the cerebral cortex.

@@

In the following section, sensory sources for gravity and self-motion perception are described in more detail. We then elaborate on how these signals may be used and integrated in the brain, and conclude the chapter with methods for studying the processing of these signals.

@@

\subsubsection{Otoliths}
The linear acceleration and gravitational forces are measured by the two otolith organs on either side of the head: the saccule and the utricule. Each otolith consists of an endolymph filled compartment containing calcium carbonate crystals known as otoconia. Due to their high inertia, these otoconia fall behind during linear acceleration, and move ahead during deceleration. In addition, they fall downward as a result of the gravitational field. The otoconia are mounted on top of a flexible polysaccharide gel, in which hair cells, similar to those found in the semicircular canals, are partly embedded \cite{goldberg2012}.

Einstein's equivalence principle states that  it is not possible to measure forces caused by linear acceleration and by gravity independently. The signal coming from the otoliths is therefore proportional to the combination of these forces \cite{fernandez1976b}, which is commonly referred to as the gravito-inertial force (GIF). Compared to the three semicircular canals, which are able to sense rotation in three dimensions due to their orthogonal organization, either side of the head only contains two otolith organs. It is still possible to sense the gravitoinertial force in three dimensions because the otolith organs are curved, and the orientation of the hair cells determined the direction of sensitivity \cite{goldberg2012}.

@@

For many actions, the brain needs to disentangle the contributions of gravity and linear acceleration to the gravitointertial force. For example, the linear vestibuloocular reflex (LVOR) that stabilizes gaze during translation should be sensitive to translation while ignoring gravity. Under normal circumstances, the brain is able to perform this task well \cite{merfeld1995}, but during extreme conditions such as in airplanes or space flight errors might occur. The somatogravic illusion \cite{glasauer1995} is an example of such error. In this illusion an airplane accelerates forward which causes an inertial force in the opposite direction. The direction of the resulting GIF will therefore be in between gravity (downward) and the intertial force (backward). The brain erroneously interprets the (majority of the) GIF as being caused by gravity, leading to the perception of "nose-up" pitch tilt.

Various theories have been proposed as to how the brain might solve the tilt-translation ambiguity. The frequency segregation hypothesis, for example, makes use of the fact that when stationary, the only force we experience is gravity. In this case, sustained (low frequency) accelerations should be  attributed to gravity, while the high frequency components should be attributed to translation \cite{paige1991, telford1997}. Other models keep track of the expected direction of gravity by integrating over the vector product of gravity and angular velocity from the semi-circular canals, and subtracting that signal from the otolith signal to obtain an estimate of linear acceleration. Because the semi-circular canals play a  crucial role in these models, they are known as multisensory integration models \cite{mayne1974,ormsby1977}. Merfeld and Zupan \citeyear{merfeld1995,merfeld2002} further refined the multisensory integration model by explicitly stating that the brain uses an (internal) model of the physical world to resolve the tilt-translation ambiguity.

The disambiguation of linear acceleration and gravity is not solely based on vestibular inputs, but also takes the  range of  possible movements into account. For example when moved on a linear sled, the probability of a participant perceiving tilt is greatly reduced \cite{wertheim2001}, indicating that cognitive processes also influence disambiguation.

@@

One complication is that the magnitude, $e$, of these eye movements not only depend on translation amplitude, $T$, but also on fixation depth, d. When fixating position $(x, d)$, the eye position at starting position is $e_0 = atan⁡(d/x) \approx d/x$. By subtracting the eye position after the translation, $e_1 = atan⁡(d/(x+T)) \approx d/(x+T)$, we obtain the eye movement amplitude, $e \approx d/T$. In Chapter 4 we explore whether the brain takes this geometry into account and compensates for fixation depth when using eye movements in self-motion perception.

\section{Reflexive uses for the self-motion}
While vection is a good cue to self-motion perception, it can cause the image on the fovea to blur which hampers visual perception. Several reflexive mechanisms attempt to keep fixation on target during self-motion. Two examples are the vestibulo-ocular reflex, the VOR, where the low-latency vestibular signal drives the reflex and the optokinetic reflex, the OKR, where optic flow acts as the source signal.

The vestibulo-ocular reflex (VOR) compensates for both rotation as well as translation, it therefore consists of two reflex arcs. The translational VOR, or TVOR, is driven by the otolith signal while the rotational VOR, or RVOR, is mainly driven by the semicircular canals. The RVOR operates during head rotation by counter rotating the eyes in the opposite direction to the head. Because its gain is about one, eye velocity is about equal to the head velocity. The TVOR operates during head translation, when the head moves orthogonally to the line of sight, simple geometry requires that the ideal TVOR response is inversely proportional to fixation distance, $e_vor = atan⁡(d/T)$. As a result, no compensatory eye movements are required when fixating targets at infinity to maintain a stable retinal image. Because the latency of the vestibular signal is much lower compared to that of the OKR, it dominates at higher frequencies \cite{schweigard1997}.

The optokinetic reflex uses the optic-flow based signal to self-motion, that is vection, to compensate for both translational and rotational movement. Because the retinal cells driving the OKR are more sensitive to slow motion, the OKR itself compensates for low frequency movements. In contrast to the vestibularly driven VOR, it is even sensitive to constant velocity stimuli \cite{soodak1988}.

In both the OKR and the VOR the eye movements made consist of two parts, the slow-phase pursuit-like movement in which the eyes are kept on target using slow eye movements and the quick-phase saccades in which the eyes are quickly moved back to their central position. The fast movements are elicited when the eyes are about to lose track of the target because of physical constraints within the oculomotor system.

@@

While reflexes by definition depend on minimally processed sensory signals, higher animal functions can rely on more elaborate processing. One such processing step is determining the underlying physical cause of a sensory signal. It has become clear that multiple sensory systems provide information about the same physical quantity. For example, both the visual and the vestibular system contribute to self-motion perception. A very naive approach to deal with these inputs would be to focus on the most reliable cue and ignore the others. A better solution, however, is to weight all available cues by their  relative reliabilities. This approach is known as statistically optimal, or Bayesian, integration, the latter referring to the use  of Bayes' inference.

@@

These reference frame transformations are not static, but depend on the position of our body. For example when converting the gaze-fixed retinal signal into a head-centred frame of reference, the position of the eyes have to be taken into account. When the position of our body changes, e.g. when we move our eyes, all coordinates remembered within the moving frame of reference also need to be updated. For example, when moving the eyes remembered retinal information needs to be shifted in the opposite direction to remain at the correct relative position. This process is known as spatial updating or spatial constancy.

@@

\subsection{Subjective body tilt}
In the subjective body tilt, or SBT, task the perceived body orientation with respect to a given body tilt angle is probed. Participants are first given a reference angle, e.g. 45 \si{\degree}, and are then rotated to an angle close to the reference angle, e.g. 46 \si{\degree} (see \figref{intro:fig4}A). They then have to indicate whether their current orientation is clockwise or counter-clockwise with respect to the reference angle. By systematically probing rotation angles around the reference angle we can obtain both the bias and uncertainty on the percept of body tilt.

Most participants are able to do this perfectly, regardless of the reference angle (see \figref{intro:fig4}B). Their uncertainty does increase as a function of reference angle though (see \figref{intro:fig4}C).
As the SBT probes body orientation, the somatosensory signal originating from the torso can be used without any reference transformation and thus provides a direct contribution. Other sensory signals such as the vestibular signal can also be used, but only after a reference frame transformation.

@@

\subsection{Subjective visual vertical}
The subjective visual vertical, or SVV, task is a similar task in which participants have to judge the orientation of a line with respect to gravity. The PSE, that is the angle at which the line is perceived to be aligned with gravity, can be found by presenting lines at different angles and asking the participant whether the line is rotated clockwise or counter-clockwise relative to gravity (see \figref{intro:fig5}A).

When seated straight this task, participants do not make any static errors and are very certain about their responses (see \figref{intro:fig5}B and C). This changes when first rotating the participant before the task. In general, the static error increases with tilt angle, the direction of the error is in the direction of the body mid-line, this effect is known as the Aubert or A-effect. At smaller angles overcompensation occurs and the static error moves in the opposite direction, that is away from the body mid-line. This latter effect is known as the E-effect and could be due to the effects of ocular counter-roll (OCR).

@@

\section{Outline of this thesis}
Multiple sensory signals provide information about both body orientation and self-motion. This thesis explores the how the brain integrates these signals to form dynamic but coherent percepts.

We start with Chapter 2 in which we investigate whether the visual, somatosensory, and vestibular cues to body orientation are integrated in a statistically optimal fashion. Because the contributions of these signals cannot be studied in isolation, we adopted an inverse approach where we assumed statistical optimality and attempted to compute the statistical properties of individual sensors. To this end, we fitted an optimal integration model to two physophyisical tasks, one gauging head-in-space orientation (SVV) and the other body-in-space orientation (SBT). Predictions based on these noise properties were consistent with previously published deficits of various patient groups, further strengthening the idea that human spatial orientation is statistically optimal.

In Chapter 3 we move on to the dynamic aspects of vestibular integration, investigating whether eye movements contribute to the perception of whole-body translation. Using a psychophysical task, we first shows that translations in which the eyes fixate a body-fixed target are perceived as shorter than those in which the eyes fixate make pursuit movements by fixating a world-fixed target. We then go on an show that this result does not depend on the presence of the fixation point by including an unconstrained eye movement condition. Because fixation distance influences the magnitude of pursuit eye movements, we fixed the distance of the fixation point in Chapter 3. To investigate whether the brain takes fixation distance into account, we compared self-motion perception during near and far world-fixed targets in Chapter 4. Our results suggest that translation is not properly accounted for, indicating that raw eye movements augment self-motion perception.

We conclude this thesis with Chapter 5, where we investigate how these self-motion signals are used to track previously seen objects across translation. First, we show that the errors made during spatial updating do not only depend on the location of the previously seen object, but also on the location of gaze. This is consistent with spatial updating in a gaze-centred reference frame. We further show that the underestimation of self-motion is a possible cause of the errors found.
